---
title: "Auckland SA2 Census Cleaning & Travel-to-Work Processing"
author: "Hinami Ariana Nishi"
format: html
execute:
  echo: true
  warning: false
  message: false
---

#### Introduction

This document outlines the full workflow used to prepare, clean, and structure the 2023 Census SA2 dataset for Auckland, with a specific focus on variables related to **main means of travel to work**.  

The workflow includes:

- loading SA2 spatial data  
- joining lookup metadata  
- renaming variables into human‑readable formats  
- clipping the dataset to the Auckland TA boundary  
- cleaning sentinel values  
- extracting travel‑to‑work variables  
- preparing additional demographic groups for later analysis  

This `.qmd` file ensures the workflow is **transparent, reproducible, and version‑controlled** for research purposes.


## Data Clean

### Load Libraries

```{r}
library(sp)
library(sf)
library(GWmodel)
library(dplyr)
library(tidyverse)
library(tmap)
library(RColorBrewer)
library(grid)
library(gridExtra)
library(terra)
library(janitor)
library(stringr)
```


### Load SA2 Data and Lookup Table

```{r}
# Name of the data: 2023 Census totals by topic for individuals by statistical area 2 – part 2
sa2_ind <- st_read("./data/SA2-individual/2023-census-totals-by-topic-for-individuals-by-statistical-a.shp")
look_up = read.csv("./data/SA2-individual/2023_census_totals_by_topic_for_individuals_by_sa2_part_2_lookup_table.csv")
```


###  Identify Travel-to-Work Variables

```{r}
# Search for columns related to travel to work
travel_vars <- look_up %>%
  filter(str_detect(Variable1, "Main means of travel to")) %>%
  select(c(Shapefile_name, Variable1, Variable1_category, Year, Field_name_alias))

#Mode Types
(unique(travel_vars$Variable1_category))

look_up[which(is.na(look_up$Year)), ]# Inspect the rows where years show NAs; we keep these

# Filter the lookup table FIRST to keep only 2023 variables and geographic metadata
look_up_2023 <- look_up %>%
  filter(Year == 2023 | is.na(Year)) # %>%# Keep variables from 2023 OR variables with NA years (geographic codes/names)
```


### Build Clean Descriptive Names

```{r}
# Build descriptive names for the 2023 dataset
look_up_cleaned_2023 <- look_up_2023 %>%
  mutate(
    descriptive_name = case_when(
      # If both main variable and category exist, combine them with the year
      !is.na(Variable1) & !is.na(Variable1_category) ~ paste(Variable1, Variable1_category, Year, sep = "_"),
      
      # Otherwise use the field name alias
      !is.na(Field_name_alias) ~ paste(Field_name_alias, Year, sep = "_"),
      
      # Fallback to the original shapefile name
      TRUE ~ Shapefile_name
    )
  ) %>%
  mutate(clean_name = make_clean_names(descriptive_name)) # (eg: "usual_residence_public_bus_2023")
# glimpse(look_up_cleaned_2023)
```



### Subset and Rename SA2 Columns

```{r}

rename_map_2023 <- setNames(look_up_cleaned_2023$clean_name, look_up_cleaned_2023$Shapefile_name)# Create a named vector mapping old names to new names: c("old_name" = "new_name")

#Explicitly define which spatial columns must be preserved
spatial_cols <- c("SA22023_V1", "SA22023__1", "SA22023__2", "AREA_SQ_KM", "LAND_AREA_SQ_KM") # del->, "Shape_Leng", "geometry"
# spatial_cols %in% names(rename_map_2023) #Sanity Check



# Subset the dataset to keep ONLY spatial columns and 2023 columns, then rename them
sa2_renamed_2023 <- sa2_ind %>%
  
  # subset columns first (dropping 2013 and 2018)
  select(any_of(c(spatial_cols, names(rename_map_2023)))) %>%
  
  # rename remaining columns in bulk
  rename_with(~ recode(., !!!rename_map_2023), .cols = any_of(names(rename_map_2023)))


```



### Clip to Auckland Boundary

```{r}
# Boundary Layer 
boundary = st_read("./data/boundary/TA-2023-clipped.shp")
akl_boundary = boundary %>%# Filter to Auckland
  filter(TA2023_V1_ == "076")

#Align projections-Ensure the Auckland boundary uses the exact same projection as your SA2 variables
akl_boundary_proj <- st_transform(akl_boundary, st_crs(sa2_renamed_2023) )
# Perform the physical clip (st_intersection)
# This removes all national data and oceanic data outside the Auckland TA boundary line
sa2_clipped <- st_intersection(sa2_renamed_2023, akl_boundary_proj)

# Clean up the clipped boundaries (This fixes the terra plotting error)
sa2_clipped_clean <- sa2_clipped %>%
  st_make_valid() %>%                      # Repair self-intersections/invalid border shapes
  st_collection_extract("POLYGON") %>%     # Drop points and lines created on the boundary edges
  st_cast("MULTIPOLYGON") %>%              # Standardize geometry types to MULTIPOLYGON
  filter(!st_is_empty(.))                  # Remove any empty geometry rows

# Convert and plot
# This will now convert and plot with terra without any geometry mismatch errors
plot(vect(sa2_clipped_clean), col = "lightblue", main = "Auckland Region Only (Clipped)")
```



### Clean Sentinel Values (-999, -997)

```{r}
#summary(sa2_clipped_clean) #we notice -999, -997 vals (typical for statsNZ dataset)
# Replace all negative values (-999, -997) with NA across all numeric columns
sa2_clipped_final <- sa2_clipped_clean %>%
  mutate(across(
    where(is.numeric), 
    ~ ifelse(. < 0, NA, .)
  ))

# sf::st_write(
#   sa2_clipped_final,
#   dsn = "./data/Saved/sa2_clipped_clean.gpkg",
#   layer = "sa2_clipped_clean",
#   delete_dsn = TRUE # Overwrites the file if it already exists
# )


# UNDERSTANDNING THE DATA - CONTEXT 
# How many neighbourhoods in my Auckland SA2?
length(unique(home_to_work$usual_residence_statistical_area_2_name_na))
```



## Understand the TRAVELLING GROUPS

```{r}
#Load Cleaned Data for Analysis
sa2_clipped_clean <- st_read("./data/Saved/sa2_clipped_clean.gpkg")
```

### Inspecting the variable 

```{r}
# Inspection
sum(str_detect(colnames(sa2_clipped_clean), "main_means_of_travel_to")) #Inspecting the num of cols of travel groups
travel_groups_inspect = sa2_clipped_clean[,which(str_detect(colnames(sa2_clipped_clean), "main_means_of_travel_to"))]
# View(travel_groups_inspect) #What these are....
sum(str_detect(colnames(travel_groups_inspect), "main_means_of_travel_to"))

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# SOME NOTES
# There are essentially 4 groups in these data:

# # 1. Trip to work
# - main_means_of_travel_to_work_by_usual_residence
# - main_means_of_travel_to_work_by_workplace_address
# 
# # Trip to education
# - main_means_of_travel_to_education_by_usual_residence_address
# - main_means_of_travel_to_education_by_education_address #What does this mean? 

# I will be focusing on the home->work dataset
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

```




### Filter to home to work travel

```{r}
# Inspect Home-to-work data set and filter
home_to_work <- sa2_clipped_clean %>%
  select(!contains(c(
    "main_means_of_travel_to_education_by_usual_residence_address", 
    "main_means_of_travel_to_education_by_education_address", 
    "main_means_of_travel_to_work_by_workplace_address"
  ))) 
```


### Rename Travel Mode Columns for convenience

```{r}
home_to_work_clean <- home_to_work %>%
  rename_with(
    ~ .x %>% 
      str_remove("^main_means_of_travel_to_work_by_usual_residence_address_") %>% 
      str_remove("_2023$") %>% 
      paste0("transportMode_", .), # Adds 'transport_' to the front
    .cols = starts_with("main_means_of_travel_to_work_by_usual_residence_address_")
  )


```



### Save Home-to-Work Dataset

```{r}
sf::st_write(
  home_to_work_clean,
  dsn = "./data/Saved/home_to_work.gpkg",
  layer = "home_to_work",
  delete_dsn = TRUE
)
```



### Inspect Additional Variable Groups

```{r}
variable_name_inspect <- function(data, str_detection) {
  matching_cols <- colnames(data)[str_detect(colnames(data), str_detection)]
  cat("Found", length(matching_cols), "columns matching pattern '", str_detection, "':\n")
  print(matching_cols)
  subset_data <- data[, matching_cols, drop = FALSE]
  return(subset_data)
}

ho_vars <- variable_name_inspect(home_to_work_clean, "individual_home_ownership_")
res1_vars <- variable_name_inspect(home_to_work_clean, "usual_residence_1_year_ago_indicator_")
res5_vars <- variable_name_inspect(home_to_work_clean, "usual_residence_5_years_ago_indicator_")
yrs_res_vars <- variable_name_inspect(home_to_work_clean, "years_at_usual_residence_")
arrival_vars <- variable_name_inspect(home_to_work_clean, "years_since_arrival_in_new_zealand_")
study_vars <- variable_name_inspect(home_to_work_clean, "study_participation_")
highest_qual_vars <- variable_name_inspect(home_to_work_clean, "highest_qualification_")
post_school_nz_vars <- variable_name_inspect(home_to_work_clean, "post_school_qualification_in_new_zealand_indicator_")
sec_qual_vars <- variable_name_inspect(home_to_work_clean, "highest_secondary_school_qualification_")
attainment_vars <- variable_name_inspect(home_to_work_clean, "post_school_qualification_level_of_attainment_")
inc_sources_vars <- variable_name_inspect(home_to_work_clean, "sources_of_personal_income_total_responses_")
total_inc_vars <- variable_name_inspect(home_to_work_clean, "total_personal_income_")
labour_status_vars <- variable_name_inspect(home_to_work_clean, "work_and_labour_force_status_")
job_search_vars <- variable_name_inspect(home_to_work_clean, "job_search_methods_total_responses_")
emp_status_vars <- variable_name_inspect(home_to_work_clean, "status_in_employment_")
unpaid_vars <- variable_name_inspect(home_to_work_clean, "unpaid_activities_total_responses_")
hours_vars <- variable_name_inspect(home_to_work_clean, "hours_worked_in_employment_per_week_")
ind_res_vars <- variable_name_inspect(home_to_work_clean, "industry_by_usual_residence_address_")
ind_work_vars <- variable_name_inspect(home_to_work_clean, "industry_by_workplace_address_")
occ_res_vars <- variable_name_inspect(home_to_work_clean, "occupation_by_usual_residence_address_")
occ_work_vars <- variable_name_inspect(home_to_work_clean, "occupation_by_workplace_address_")
transport_vars <- variable_name_inspect(home_to_work_clean, "transportMode_")
sector_vars <- variable_name_inspect(home_to_work_clean, "sector_of_ownership_")
source_vars <- variable_name_inspect(home_to_work_clean, "individual_unit_data_source_")
```



### 18. Next Steps

You can now:

- map travel‑to‑work patterns  
- analyse mode share differences across SA2  
- integrate vehicle ride‑time (Google Directions API)  
- build OD matrices  
- model accessibility or congestion impacts  

This cleaned dataset is ready for spatial analysis, modelling, and visualisation.
